Extended Data

### 2022-03-11 doc/manuscript/Dow_manuscript_Nature.Rmd
restructuring discussion

<!--1. summary-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].

Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].

Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016;@fu_maximum_2019]. 
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
We show that the extra C fixation in years with warm springs does not substantially augment stem growth in the current of following year (`r EDF_tree_rings_TMAX_extended`), but it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

<!-- The following two paragraphs--and two final sentences in the paragraph above-- need work (assigned to KAT). It may make most sense to condense into a single paragraph making the point that while we don't know where the C goes, we can rule out a substantial, sustained boost to the long-term sink of woody productivity-->

~~~*points to make*:

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
- our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs]. Specifically, [explain]. The implication is that while elements of global change such as warming spring temperatures or elevated CO~2~[@refs] may enhance GPP, woody growth remains constrained.
- we do not know where the extra C is going. If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs]. We can also look to studies documenting enhanced GPP with no enhancement of growth. In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
~~~

One possibility is that extra photosynthate in years with warm springs may be allocated to woody growth without affecting diameter growth in the current year. 
It is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a thicker layer of higher-density latewood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Thus, warm springs are unlikely to provide substantial, sustained C sinks under warming spring temperatures.

A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
<!--gao_earlier_2022 demonstrates the same for some other regions.-->
<!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.

>>>>

<!--summary-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].

<!--expectation of extra C, not going to ANPP_woody (sink limitation)-->
Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022;@tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016;@fu_maximum_2019]. 
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018;@crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
We show that the extra C fixation in years with warm springs does not substantially augment stem diameter growth in the current or following year (`r EDF_tree_rings_TMAX_extended`).

Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016 and refs therein].  
Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[@tardif_influence_2006;@roibu_climatic_2020], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
*Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that $ANPP_{woody}$ is limited by sink factors, such that the expected increase photosynthate availability[@refs] (observed at Harvard Forest[@finzi_carbon_2020_but_verify]) does not enhance growth.*
*Indeed, such a phenomenon has been observed when GPP has been increased through lengthening of the growing season[@refs] or elevated CO~2~[@refs].*

<!--on the fate of extra C-->
*notes:*

- our study adds to a growing body of literature demonstrating a sink limitation of stem growth and $ANPP_{woody}$ [@refs]. Specifically, [explain]. The implication is that while elements of global change such as warming spring temperatures or elevated CO~2~[@refs] may enhance GPP, woody growth remains constrained.
- we do not know where the extra C is going. If we look at instances where growth is more strongly limited than GPP, we see that NSCs can build up[@refs], and this build up can feed back to reduce photosynthesis[@refs]. We can also look to studies documenting enhanced GPP with no enhancement of growth. In a mature eucalypt forest exposed to elevated CO2, enhanced GPP did not result in higher growth, but rather was released back to the atmosphere on relatively short time scales through enchanced respiration[@jiang_fate_2020]
- lessons learned from the more intense study of CO2 fertilization imply that stimulation of ANPP_woody is likely to be less than that of GPP

*old content:*
A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.

<!--... short vs long-term effects-->

*notes:*

- we show lack of effect on interannual scale 
- feedbacks such as greater nutrient acquisition could change this.

*old content:*
Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->

- Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022]. 
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018;@lian_summer_2022]
- our analysis isn't set up for detecting interaction. But, sites are fairly mesic. Mechanisms remain to be understood, but the net result is the same.

<!--conclusions-->

*old content:*
As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier; however, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. 1) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback that is anticipated in most climate forecasting models[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
<!--gao_earlier_2022 demonstrates the same for some other regions.-->
<!-- cite @gao_earlier_2022, noting that their result may be driven by conifers-->
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. 3), which constitutes a positive feedback to climate change.
